Categorical and gradient aspects of wordlikeness\thanks{Thanks to Gene Buckley, Charles Yang, and audiences at NELS 43 for helpful comments.}

### 2013-01-28 2.tex
Revisions! Lots of 'em.

One of the major  for the development of stochastic models o
motivations for the 
Recent developments in the theory of phonotactic knowledge are motivated by the claim that speakers' wordlikeness intuitions are inherently \emph{gradient}, i.e., that such intuitions are more granular than a mere binary contrast between ``possible'' and ``impossible'' words.

@@

\ldots{}when judgements are elicited in a controlled fashion from speakers, they always emerge as gradient, including all intermediate values. \citep[371]{Shademan2006} 
\end{quote}

This is not a novel claim. Early generative discussions of wordlikeness \citep[e.g.,][]{Chomsky1965,Halle1962} are best remembered for the famous examples [blɪk] and [bnɪk], the former representing a ``possible word'' of English and the latter representing an ``impossible word''. 
A naïve account of this contrast follows from the assumption that segments must be parsed into syllables or subject to further phonological repair (e.g., \citealt[10f.]{Hooper1973}, \citealt[57f.]{Kahn1976}, \citealt{Ito1989a}, \citealt[19f.]{Wolf2009}). 
Unlike some languages (e.g., Morroccan Arabic: \emph{bniqa} `closet'), English does not permit stop-nasal onsets like [bn], so the latter nonce word cannot surface as such. 
In other words, [bnɪk] is an impossible surface representation in English. 
However, in \emph{The Sound Pattern of English}, \citet{SPE} introduce a third nonce word, [bznk], and claim that last form is in some sense even less English-like than [bnɪk].\footnote{
    That wordlikeness judgements depend on language-specific knowledge is apparent given that [bznk] is not impossible in all languages: Imdlawn Tashlhiyt Berber permits whole words consisting of a stop-fricative-nasal-stop sequence (e.g., [tzmt] `it is stifling'; \citealt[112]{Dell1985}). 
    This suggests that wordlikeness judgements depend on language-specific knowledge.
    It has more recently been suggested some contrasts between unattested clusters can be viewed as cross-linguistic implicational universals \citep[e.g.,][]{Berent2007a,Berent2008a,Berent2009,Berent2007b}, so that, for instance, the acceptance of [bznk] might imply acceptance of [bnɪk]. 
    If this is correct, it undermines \citeauthor{SPE}'s position.}
From this they too conclude that wordlikeness intuitions are gradient.

\noindent
This brings the theory of wordlikeness in line with the view of syntactic grammaticality presented by foundational documents like \emph{The Logical Structure of Linguistic Theory} \citep{LSLT} and \emph{Aspects of the Theory of Syntax} \citep{ASPECTS}, which posits multiple degrees of ungrammaticality arising from different types of syntactic violations.
\citet{BARRIERS} and \citet{Huang1982} propose further elaborations of this type of theory (see \citealt[43f.]{Schutze1996} for a critique).

It has been recognized, however, that the naïve account of the [blɪk] and [bnɪk] cannot easily be extended to account for these ``numerous other distinctions''.

\begin{quote}
A defect of current grammatical accounts of phonotactics is that they render simple up-or-down decisions concerning well-formedness and cannot account for gradient judgements. \citep[371]{Shademan2006}
\end{quote}

Scientific observations do not necessarily arrive at the appropriate granularity for analysis.
Since the granularity of measurement and analysis is determined by prior hypotheses about the nature of the system under study, disputes are inevitable as scientific paradigms evolve. 
This chapter argues that there are both theoretical and empirical reasons to doubt the implicit hypothesis linking scalar wordlikeness judgements and a gradient model of phonotactics. 
First, intermediate judgements are characteristic of all gradient rating tasks, and therefore these judgements are irrelevant to the question of whether well-formedness is categorical or gradient.
Secondly, simple baselines better account for gradient well-formedness judgements than current computational models of phonotactic knowledge, suggesting that the gradience observed in these tasks do not derive from grammatical mechanisms.

%are ``\ldots{}under the obligation to say why \emph{what it takes to be} data relevant to the confirmation of its theories \emph{are} data relevant to the confirmation of its theories'' \citep[200]{Fodor1981b}.
%The naïve account of the [blɪk]/[bnɪk] contrast presented above is a special case of this

@@

The aforementioned discussions of gradient aspects of wordlikeness judgements take for granted that intermediate judgements require a model of gradient grammaticality. 
The general position that such data can be taken at face value is known in the cognitive sciences as \emph{naïve realism} \citep{Fodor1981a}, and is not uncontroversial. 
With respect to wordlikeness, there are numerous reasons to question the wisdom of this position.

@@

The results of controlled experiments are often biased by subtle details that seem orthogonal to the task: for instance, certain types of duration judgements are systematically biased by consumption of caffeine \citep{Gruber2005}. 
It should come as no surprise, then, that a highly salient aspect of a judgement task, the scale used for responses, also influences the results obtained.
\citet*{Armstrong1983} argue that the use of many-valued scales may cause intermediate ratings, and should be interpreted as a task effect.

@@

It might be said that these results reveal something about the representation of odd numbers, but no scientist of mathematical cognition has risen to the challenge of constructing a theory that might account for the fact that 447 is rated more odd than 3. 
\citeauthor{Armstrong1983} anticipate this objection.

@@

It is possible to draw an analogy: can it be the case that [st] is a significantly ``better'' onset than [bl] (a prediction of the wordlikeness model proposed by \citealt{Albright2009a}, for instance), but ensure that both are treated the same with respect to syllabification? 
Current research on gradient well-formedness is concerned with specifying the component by which a scalar value is assigned to linguistic structures. 
This is only one part of any model of gradient grammaticality, however.
Further assumptions are necessary, as follows.
When presented with a linguistic item (of whatever type) in a well-formedness task, the grammar assigns a parse.
This grammar must be able to parse an enormous range of linguistic structures, including many which it cannot be permitted to generate; independent perception and production grammars may be necessary. 
A scalar value is then assigned to this parse. 
Then, speakers consciously access this scalar value, then transform it in accordance with the numerical scale chosen by the experimenter.

Each step of this procedure is dubious, however.
First, speakers have difficulty perceiving \citep{Berent2007a,Brown1956,Dupoux1999,Kabak2007a} and producing \citep{Davidson2005,Davidson2006a,Davidson2006b,Davidson2010,GallagherInPress,Rose2007,Vitevitch1998,Vitevitch2005} phonotactically illicit non-words, suggesting that speakers' ability to faithfully parse illicit representations is at best quite limited.
Secondly, the computation of a scalar value serves no further purpose than to provide for gradient well-formedness judgements; even if this is the result of probabilistic parsing \citep[e.g.,][]{Coleman1997}, 
it need not be computed overtly.\footnote{
    An objection might be made here on evolutionary grounds. 
    It is still quite mysterious why the human language endowment includes constraints on pronominal binding, for instance, but these constraints are implicated in everyday language use.
    Far more bizarre is the suggestion that the linguistic endowment includes mechanisms only used in certain experimental tasks.}
Next, speakers must be able to consciously access and report the magnitude of this value (it must be \emph{cognitively penetrable} in the sense of \citealt{Pylyshyn1984}), an ability which is limited in many other domains.
Finally, \citet{Sprouse2011} argues that speakers do not (or cannot) scale well-formedness values to satisfy the assumptions of magnitude estimation, a type of gradient rating task.
%\citep{Bard1996}

It is informative to compare this baroque architecture to a model of what is necessary to make a binary well-formedness judgement. When presented with a linguistic item in a judgement task, the grammar attempts to assign a parse. Speakers then access whether or not parsing was successful. 
There are reasons to think that parsing of ungrammatical structures does in fact result in a ``crash'': whereas syntactic priming increases the acceptability of grammatical structures \citep{Luka2005}, ungrammatical structures show no priming effects \citep{Sprouse2007b}. 
As priming of linguistic structures is thought to implicate shared representations in memory, ungrammatical structures may not be stored in linguistic memory. 
%This cannot be reduced to the fact that ungrammatical structures fail to denote, since well-formed but non-existent structures do produce facilitory priming \citep[e.g.,][]{Longtin2005}. 
The fact that requests for repetition and clarification are ubiquituous in spontaneous speech illustrates further that speakers are frequently aware when parsing has failed.
Consequently, a great deal of evidence is needed to reject this simple model in favor of the gradient grammaticality architecture.

@@

\citet{Hayes2000} argues that it is ``uninsightful'' to attribute gradience to task effects, insofar as these effects call upon grammatical representations.

@@

\noindent
This is indisputable. 
However, there have been no serious attempts to evaluate categorical and gradient models of wordlikeness on an equal footing.
In light of the complexities of gradient models, such an evaluation requires strong quantitative evidence for the superiority of gradient grammatical models.
The evaluation below represents a first attempt to fill this gap.

It is not that categorical models have been ignored by the literature on wordlikeness modeling, but rather that they have not been compared. 
\citet{Frisch2000} and \citet{Vitevitch1997} find that speakers' wordlikeness ratings of multisyllabic words are correlated with a probabilistic measure of the well-formedness of the constituent syllables. 
Unfortunately, no attempt is made to control for the well-formedness of syllable contact clusters in these words: some of the stimuli have medial consonant clusters containing both voiced and voiceless obstruents (e.g., [ɡaɪbsaɪk]), something which is exceptionally rare in English simplex words.
Similarly, \citet{Hayes2008a}, who compare their gradient model of wordlikeness against a set of phonotactic constraints proposed by \citet{Clements1983}, first transform these constraints, many of which are exceptionless, into probabilities. 
While this is consistent with their claim, that ``the ability to model gradient intuitions [are] an important criterion for evaluating phonotactic models'' \citep[382]{Hayes2008a}, this principle would preclude any attempt to test the hypotheses that underlies it.

%(see \citealt{Sproat1993} and \citealt{Wrench2003} for articulatory investigations).

@@

It is unknown whether the intermediate ratings in gradient wordlikeness tasks are reliably predicted by computational models that have been proposed. 
If some model poorly accounts for these intermediate ratings, it may be the case that the model is improperly specified, or it may indicate that a considerable amount of variance in wordlikeness ratings is not a product of the grammatical system per se. 
It seems particularly plausible that speakers might differentiate, in a regular fashion, between different types of ``impossible'' words, as proposed by \citet{SPE}. 
There are also claims that speakers distinguish between different types of ``possible'' words, so that, for instance, [stɪn] \emph{stin} is rated more English-like than [blɪn] \emph{blin} \citep{Albright2009a}. 
Alternatively, wordlikeness judgements could be effectively modeled with a gross contrast between possible and impossible words; even then, a gradient model might show a residual correlation.
All of these possibilities are considered here.

@@

This evaluation uses a large sample of three previously published studies on English wordlikeness comprising 125 subjects and 187 items. Two criteria were used to select these three studies. First, it was required that the stimuli be presented aurally so as to eliminate any possibility of orthographic effects \citep[e.g.,][]{Berent2001b,Berent2008b}. Secondly, the data must be sufficiently ``phonotactically diverse'': that is, it must include both items like \emph{blick} and \emph{bnick}. This excludes studies like that of \citet{Bailey2001}, in which few if any items contain gross phonotactic violations of the type represented by \emph{bnick}, since this is likely to minimize the coverage of any grammatical model. The data is summarized in Table \ref{counts}.

@@

Models are evaluated by comparing their scores to the average rating of each word using four correlation statistics. Each of these range between $[-1, 1]$, where $1$ indicates a perfect positive correlation and $-1$ denotes a perfect negative correlation. \citet{Hayes2008a} evaluate their model using the Pearson (``product-moment'') $r$, a parametric correlation measure.  \citet{Stevens1946} argues that statistics of this type are inappropriate for analysis of Likert scale data, like those used by \citet{Albright2007} and \citet{Albright2003b}.
The reason is that the Pearson $r$ makes a \emph{linearity assumption}. That is, it assumes that nonce words rated ``1'' and  ``3'', for instance, are just as different as are those rated ``4'' and ``6''. A weaker assumption, more appropriate for Likert scale data, is the \emph{monotonicity assumption}: that ``1'' is less English-like than ``3'', which is less English-like than ``4'', and so on. However, it also has been claimed that $r$ is particularly robust to violations of the linearity assumption \citep[e.g.,][]{Havlicek1976}. Pearson $r$ is reported here, but no stance is taken on its appropriateness for this data.
%\citet{Jamieson2004}
\citeauthor{Hayes2008a} also report Spearman $\rho$; this statistic requires only the weaker assumption of monotonicity, but it is difficult to give a simple interpretation to the coefficient.

Two other non-parametric statistics the Goodman-Kruskal $\gamma$ and the Kendall $\tau_b$ are much easier to interpret, as follows \citep{Noether1981}. 
These statistics are computed by comparing every model score/wordlikeness rating pair to every other: a comparison is counted as \emph{concordant} if the greater of the two model scores is the one associated with the greater of the two wordlikeness ratings (that is, the model ranks these two nonce words in accordance with speakers' ratings), and as \emph{discordant} otherwise. 
These two statistics differ only in the treatment of ``ties'', pairs where either the model score or wordlikeness rating are identical. 
For $\gamma$, ties are ignored, and the coefficient is

@@

Operationalizing ``phonotactic violation'' is somewhat more difficult. The simplest possible mechanism is chosen here: an onset or rime is identified as well-formed if it occurs with non-zero frequency in a representative sample, and is identified as ill-formed otherwise. The sample is derived from those entries of the CMU pronunciation dictionary which occur at least once per million words in the SUBTLEX-US frequency norms, the latter thought to be particularly strongly correlated with behavioral measures \citep{Brysbaert2009}. These pronunciations are then syllabified, and individual syllables parsed into onset and rime, according to a process described in detail in Appendix \ref{syllabification}. This is not to imply that all unattested onsets or rimes should be regarded as ill-formed, or that all onsets or rimes with non-zero frequency in this data are well-formed. For instance, \citet{Albright2009a} judges [dɹɛsp] \emph{dresp} to be phonotactically well-formed, despite the total lack of [ɛsp] rimes in English. Similar observations have been made concerning English onsets \citep[e.g.,][]{Cairns1972,Moreton2002}. In the other direction, there are precedents for labeling certain attested words as phonotactically ``peripheral'' (e.g., \citealt{Myers1987}, \citealt{Borowsky1989}), as lexical exceptions to language-specific prosodic constraints, without labeling these words as absolutely ungrammatical. Neither of these dissociations between attestation and well-formedness are implemented here, however.

In Figure \ref{boxplot}, wordlikeness ratings from the three studies are plotted according to this gross contrast. While there are a considerable number of outliers, there can be little doubt that the contrast is reflected in wordlikeness judgements in all three studies.

@@

A wide variety of lexical similarity measures were considered, including a variant of the Generalized Neighborhood model \citep{Bailey2001}, PLD20 \citep{Suarez2011}, and a set of measures provided by the Irvine Phonotactic Calculator \citep{Vaden2009}. The most reliable measure is also the most venerable measure of lexical similarity: Coltheart's $N$ \citep{Coltheart1977}, which is defined as the number of words in some representative sample which can be changed into a target nonce word by a single insertion, deletion, or substitution of a phone. \citet{Greenberg1964} find a correlation between wordlikeness ratings and a variant of this measure which only counts words differing by a single substitution. This is plotted against ratings from the three studies in Figure \ref{neighborhood}, where it can be seen that it accounts for much of the variance in ratings.

@@

\begin{figure}[t]
\centering
\includegraphics{neighborhood.pdf}
\caption{Correlation between Coltheart's $N$ and item-averaged wordlikeness ratings}
\label{neighborhood}
\end{figure}

While there is nothing inherently ``phonotactic'' about Coltheart's $N$, it indirectly incorporates much of the information present in the gross phonotactic baseline. 
Consider [blɪk]: since there is nothing marked about any part of this nonce word, a ``neighbor'' might be found by modifying any phone: e.g., \emph{click}, \emph{brick}, \emph{bloke}, \emph{bliss}. 
However, since [bn] onsets are unattested in English, a neighbor of [bnɪk] must somehow modify this cluster: this leaves only \emph{brick} and \emph{nick}. 
\citet{Bailey2001} and \citet{Frauenfelder1993} note that neighborhood density is also strongly correlated with measures like bigram probability, but it is argued elsewhere that phonotactic measures and neighborhood density have distinct effects \citep[e.g.,][]{Berent2003,Pitt1998b,Vitevitch1998,Vitevitch1999a}.

@@

Faciliatory effects of bigram probabilities (i.e., shorter latencies) are reported for other nonce word tasks conducted with adults, including single-word shadowing \citep{Vitevitch1997,Vitevitch1998}, same/different judgements \citep{Lipinski2005,Luce2001,Vitevitch1999a,Vitevitch2005}, and lexical decision \citep{Pylkkanen2002a}. \citet{Albright2009a} applies them as a model of wordlikeness judgements. The bigram probability of a sequence $ijk$, for instance, is:

@@

\citet{Albright2009a} compares two variants of this model, the first operating over segments, the second over sets of features. Unfortunately, the latter model is not described in sufficient detail to allow it to be implemented directly, and there is no publicly available implementation. However, \citeauthor{Albright2009a}'s evaluation, which includes the \citet{Scholes1966} and \citet{Albright2003b} data, finds an advantage for segmental bigrams. In implementing this model, it was found that a slight improvement could be made by preventing any phone-to-phone transition from having zero probability. This is accomplished by adding 1 to the count of every transition, a technique used in natural language processing and known as Laplace, or ``add one'', smoothing. As can be seen in Table \ref{bigramcomparison}, this results in a slight increase in the correlation between the scores from this model and wellformedness ratings. This smoothed segmental bigram score is adopted below. In Figure \ref{bigram}, it is plotted against wordlikeness ratings from the three studies.

>>>>

Much of the recent work in phonotactic theory has attempted to incorporate the intuition that phonotactic wellformedness is not an ``all-or-nothing'' matter.
Rather, it is alleged, wellformedness judgements have more fidelity than is implied by a simple contrast between ``possible'' and ``impossible'', and therefore must be measured and studied at a greater degree of granularity.

This is hardly a novel claim, though it has taken on greater import with the emergence of computational models of wordlikeness.
Early generative discussions of wordlikeness \citep[e.g.,][]{Chomsky1965,Halle1962} are best remembered for the famous examples [blɪk] and [bnɪk], the former representing a ``possible word'' of English and the latter representing an ``impossible word''. 
A naïve account of this contrast would be to derive it from the assumption that segments must be parsed into syllables or subject to further phonological repair (e.g., \citealt[10f.]{Hooper1973}, \citealt[57f.]{Kahn1976}, \citealt{Ito1989a}, \citealt[19f.]{Wolf2009}).
Unlike some languages (e.g., Morroccan Arabic: \emph{bniqa} `closet'), English does not permit stop-nasal onsets like [bn], so the latter nonce word cannot surface as such.
In other words, [bnɪk] is an impossible surface representation in English.
However, in \emph{The Sound Pattern of English} (henceforth, \emph{SPE}), \citet{SPE} introduce a third nonce word, [bznk], constructed so as to be even less English-like than [bnɪk].\footnote{
    That wordlikeness judgements depend on language-specific knowledge is apparent given that [bznk] is not impossible in all languages: Imdlawn Tashlhiyt Berber permits whole words consisting of a stop-fricative-nasal-stop sequence (e.g., [tzmt] `it is stifling'; \citealt[112]{Dell1985}). 
    This is clear evidence for the assumption that wordlikeness depends on language-specific knowledge.}
This leads \citeauthor{SPE} to conclude that wordlikeness intuitions are gradient.

This brings the theory of wordlikeness in line with the view of syntactic grammaticality presented by foundational documents like \emph{The Logical Structure of Linguistic Theory} \citep{LSLT} and \emph{Aspects of the Theory of Syntax} \citep{ASPECTS}, and reflexes can be found in later work, such as the proposals of \citet{BARRIERS} and \citet{Huang1982}; see \citealt[43f.]{Schutze1996} for a critique.

\citeauthor{SPE}'s claim about the gradient nature of phonotactic wellformedness does not seem to have had much of an impact on practices of the time---as can be seen from discussion in the previous chapter, contemporary critiques focused on other elements of the \emph{SPE} phonotactic theory---but reflexes can once again be found in later work: for instance, \citet{Borowsky1989}, \citet[50f.]{Clements1983}, and \citet{Myers1987} all assume a contrast between ``peripheral'' and absolutely ungrammatical sound sequences.

Recent discussions of gradient grammaticality in wordlikeness attempt to present experimental support for \citeauthor{SPE}'s intuitions:

@@

A defect of current grammatical accounts of phonotactics is that they render simple up-or-down decisions concerning well-formedness and cannot account for gradient judgements. But when judgements are elicited in a controlled fashion from speakers, they always emerge as gradient, including all intermediate values. \citep[371]{Shademan2006}
\end{quote}

If the presence of intermediate values in wordlikeness tasks is evidence for the gradient nature of phonotactic wellformedness, then it follows that wordlikeness intuitions should be measured and modeled with a high degree of granularity.
For instance, this would be strong evidence against the naïve account of the [blɪk]-[bnɪk] contrast just alluded to, since it cannot easily be extended to account for ``numerous other distinctions''.
However, this chapter argues that there are both theoretical and empirical reasons to doubt the implicit hypothesis linking scalar wordlikeness ratings and gradient wellformedness.
First, intermediate ratings are characteristic of all gradient rating tasks, and therefore are irrelevant to the question of whether the internal phonotactic system is categorical or gradient.
Secondly, simple baselines better account for gradient well-formedness judgements than current computational models of phonotactic knowledge, suggesting that the gradience observed in these tasks do not derive from known grammatical mechanisms.

%are ``\ldots{}under the obligation to say why \emph{what it takes to be} data relevant to the confirmation of its theories \emph{are} data relevant to the confirmation of its theories'' \citep[200]{Fodor1981b}.

@@

The aforementioned discussions of gradient aspects of wordlikeness judgements takes for granted that intermediate ratings are the product of an internal system of gradient grammaticality.
This view is itself an instance of what is known as \emph{common-sense} or \emph{naïve realism}; in the cognitive sciences, this often takes the form of the assumption that experimental data can be taken at face value, without mediation from other sources of information.
However, there are several arguments for \emph{a priori} skepticism about the (a) linguistic abilities required for reporting gradient grammaticality judgements, (b) intermediate acceptability ratings as evidence for gradient grammaticality, and (c) the total lack of previous attempts to consider categorical models of wellformedness.

Current research into gradient wellformedness is concerned with specifying a function from sound sequences to scalar judgements, and thus describes the wellformedness system at a level of some abstraction, corresponding roughly to what \citet{Marr1982} calls the ``computational'' level of description.
This is only one part of any model of gradient grammaticality, however; further assumptions are needed to articulate the internal representations and algorithms by which this function computes.

First, consider the architecture which is implied by any system of gradient grammaticality.
It is essential that a system of gradient grammaticality have access to a relatively faithful representation of stimuli in a wellformedness task, and therefore it must be able to parse an enormous range of linguistic structures, including many which are not generated by the grammar itself; independent perception and production grammars may be necessary. 
A scalar value, representing wellformedness, must then be assigned to this parse.
To report wellformedness, the speaker must transform this scalar value in accordance with the numerical scale chosen by the experimenter.

Each step of this procedure merits scrutiny, however.
First, speakers have difficulty perceiving \citep{Berent2007a,Brown1956,Dupoux1999,Kabak2007a} and producing \citep{Davidson2005,Davidson2006a,Davidson2006b,Davidson2010,GallagherInPress,Rose2007,Vitevitch1998,Vitevitch2005} phonotactically illicit non-words, suggesting that speakers' ability to faithfully parse illicit representations is at best quite limited.
Secondly, the computation of a scalar value serves no further purpose than to provide for gradient well-formedness judgements, so an objection might be made here on evolutionary grounds.
It is quite mysterious why the human language endowment includes constraints on pronominal binding, for instance, but no one can doubt that these constraints are implicated in everyday language use; far more bizarre is the suggestion that the linguistic endowment includes mechanisms only implicated in certain experimental tasks.
Next, speakers must be able to consciously access and report the magnitude of this value (it must be \emph{cognitively penetrable} in the sense of \citealt{Pylyshyn1984}), an ability which is limited in many other domains.
Finally, there is some evidence that speakers do not (or perhaps, cannot) respect the numerical scales chosen by experimenters \citep{Sprouse2011}.

It is informative to compare this baroque model to the architecture implied by a binary well-formedness judgement.
When presented with a linguistic item in a judgement task, the grammar attempts to assign a parse. 
Speakers then access whether or not parsing was successful.
There are reasons to think that parsing of ungrammatical structures does in fact result in a ``crash'': whereas syntactic priming increases the acceptability of grammatical structures \citep{Luka2005}, ungrammatical structures show no priming effects \citep{Sprouse2007b}.
As priming of linguistic structures is thought to implicate shared representations in memory, this suggests different memory mechanisms for grammatical and ungrammatical linguistic objects.
%This cannot be reduced to the fact that ungrammatical structures fail to denote, since well-formed but non-existent structures do produce facilitory priming \citep[e.g.,][]{Longtin2005}. 
Finally, the fact that requests for repetition and clarification are ubiquituous in spontaneous speech illustrates further that speakers are frequently aware when parsing has failed.
Consequently, a great deal of evidence is needed to reject this simple model in favor of the gradient grammaticality architecture.

@@

The results of controlled experiments are often biased by subtle details that seem orthogonal to the task: for instance, certain types of duration judgements are systematically biased by consumption of caffeine \citep{Gruber2005}.
It should come as no surprise, then, that a highly salient aspect of a judgement task, the scale used for responses, also influences the results obtained.
\citet*{Armstrong1983} argue that rating tasks using many-valued scales may induce intermediate ratings as a task effect.

@@

It might be said that these results reveal something about the representation of odd numbers.
\citeauthor{Armstrong1983} anticipate this objection.

@@

No scientist has risen to the challenge of constructing a theory that might account for the fact that 447 is rated more odd than 3, and as \citeauthor{Armstrong1983} suggest, it is unclear whether such a theory could preserve more central facts about oddness.
Here it is possible to draw an analogy to phonotactic theory.
According to current orthodoxy, the wellformedness of a sequence is closely related to its type frequency (i.e., frequency in the lexicon).
Is it then the case that [bl] is a significantly ``better'' onset than [kl], simply because the former is approximately twice as frequent, and if so, is it possible to also ensure that these sequences are treated the same with respect to syllabification?

@@

\citet{Hayes2000} argues that it is ``uninsightful'' to attribute gradience to task effects, insofar as these effects implicate grammatical representations.

@@

\noindent
The logic of this implication is indisputable.
However, there is little empirical support for the claims that absolute and gradient well-formedness derive from similar principles; indeed, there have been no prior attempts to evaluate categorical and gradient models of wordlikeness on an equal footing.
In light of the complexities of gradient models, such an evaluation requires strong quantitative evidence for the superiority of gradient grammatical models.
The evaluation below represents a first attempt to fill this gap.
%FIXME clements and keyser

It is not that categorical models have been ignored by the literature on wordlikeness modeling, but rather that they have not been compared.
\citet{Frisch2000} and \citet{Vitevitch1997} find that speakers' wordlikeness ratings of multisyllabic words are correlated with a probabilistic measure of the well-formedness of the constituent syllables. 
Unfortunately, no attempt is made to control for the well-formedness of syllable contact clusters in these words: some of the stimuli have medial consonant clusters containing both voiced and voiceless obstruents (e.g., [ɡaɪbsaɪk]), something which is exceptionally rare in English simplex words (see \S\ref{ova}).
Similarly, \citet{Hayes2008a}, who compare their gradient model of wordlikeness against a set of English phonotactic constraints proposed by \citet{Clements1983}, first transform these constraints, several of which are without exception, into probabilities.
While this is consistent with their claim, that ``the ability to model gradient intuitions [is] an important criterion for evaluating phonotactic models'' \citep[382]{Hayes2008a}, little insight can be gained by annotating an exceptionless rule ``$p = 1.0$''. 
\citeauthor{Hayes2008a}'s principle precludes any attempt to test the hypotheses that underlies it, and therefore must be rejected.

% FIXME make this work: 
%(see \citealt{Sproat1993} and \citealt{Wrench2003} for articulatory investigations).

@@

As \citet{Newmeyer2007} writes ``the idea that categoricity is not represented in the data itself is a truism. Whether distinctions of grammaticality (as opposed to acceptability) are binary is a difficult question.'' (398)
The mere presence of gradience in judgements cannot falsify the claim of gradient grammaticality; another method is needed to evaluate this claim.
As a first step towards a falsifiable theory of gradient wordlikeness, the remainder of this chapter considers intermediate ratings in gradient wordlikeness tasks are reliably predicted by the computational models that have been proposed.
If a model is incapable of accounting for intermediate ratings, there are two possibilities: either the model itself is improperly specified and therefore at fault, or the inputs and outputs of the model are unrelated to the actual causes of the intermediate ratings, \emph{contra} \citet{Hayes2000}.

It is plausible that speakers might differentiate, in a regular fashion, between different types of ``impossible'' words, and a gradient model should reliably predict the distinctions that speakers make.
There are also claims that speakers distinguish between different types of ``possible'' words, so that, for instance, [stɪn] \emph{stin} is rated more English-like than [blɪn] \emph{blin} \citep[e.g.,][]{Albright2009a}, because the former onset is more frequent in the English lexicon.
Even if wordlikeness judgements can be effectively modeled with a gross contrast between possible and impossible words, a gradient model might show a correlation with the residual ratings.
All of these possibilities are considered below.

@@

This evaluation uses a large sample of three previously published studies on English wordlikeness comprising 125 subjects and 187 items.
Two criteria were used to select these three studies.
First, the stimuli must be presented aurally so as to eliminate any possibility of orthographic effects \citep[e.g.,][]{Berent2001b,Berent2008b}. 
Secondly, the data must be sufficiently ``phonotactically diverse'': that is, it must include both items like \emph{blick} and \emph{bnick}. 
This excludes studies like that of \citet{Bailey2001}, in which few if any items contain gross phonotactic violations of the type represented by \emph{bnick}.
In the absence of phonotactic violences, little variance in wordlikeness ratings can be attributed to phonotactic wellformedness, making it difficult to determine the coverage of gradient wellformedness models.
The data used here is summarized in Table \ref{counts}.

@@

Models are evaluated by comparing their scores to the average rating of each word using four correlation statistics. 
Each of these range between $[-1, 1]$, where $1$ indicates a perfect positive correlation and $-1$ denotes a perfect negative correlation. 
\citet{Hayes2008a} evaluate their model using the Pearson (``product-moment'') $r$, a parametric correlation measure.
It has long been argued \citep[e.g.,][]{Stevens1946} that parametric statistics are inappropriate for analysis of Likert scale data, like those used by \citet{Albright2007} and \citet{Albright2003b}, since the Pearson $r$ makes a \emph{linearity assumption}.
That is, it assumes that nonce words rated ``1'' and  ``3'', for instance, are just as different as are those rated ``4'' and ``6''. 
A weaker assumption, more appropriate for Likert scale data, is the \emph{monotonicity assumption}: that ``1'' is less English-like than ``3'', which is less English-like than ``4'', and so on.
However, it also has been claimed that $r$ is particularly robust to violations of the linearity assumption \citep[e.g.,][]{Havlicek1976}.
Pearson $r$ is reported here, but this should not be taken to imply an endorsement of its use for Likert scale data.

\citeauthor{Hayes2008a} also report Spearman $\rho$; this statistic requires only the weaker assumption of monotonicity, but it is difficult to give a simple interpretation to the coefficient.
Two other non-parametric statistics, the Goodman-Kruskal $\gamma$ and the Kendall $\tau_b$ are much easier to interpret, as follows \citep{Noether1981}. 
These statistics are computed by comparing every model score/wordlikeness rating pair to every other: a comparison is counted as \emph{concordant} if the greater of the two model scores is the one associated with the greater of the two wordlikeness ratings (that is, the model ranks these two nonce words in accordance with speakers' ratings), and as \emph{discordant} otherwise.
These two statistics differ only in the treatment of ``ties'', pairs where either the model score or wordlikeness rating are identical.
For $\gamma$, ties are ignored, and the coefficient is

@@

Operationalizing ``phonotactic violation'' is somewhat more difficult.
The simplest possible mechanism is chosen here: an onset or rime is identified as well-formed if it occurs with non-zero frequency in a representative sample, and is identified as ill-formed otherwise.
This is not to imply that all unattested onsets or rimes should be regarded as ill-formed, or that all onsets or rimes with non-zero frequency in this data are well-formed.
For instance, \citet{Albright2009a} judges [dɹɛsp] \emph{dresp} to be phonotactically well-formed, despite the total lack of [ɛsp] rimes in English; similar observations have been made concerning English onsets \citep[e.g.,][]{Cairns1972,Moreton2002}.

The representative sample used to define the phonotactic baseline is derived from those entries of the CMU pronunciation dictionary which occur at least once per million words in the SUBTLEX-US frequency norms; these norms are known to be particularly strongly correlated with other behavioral measures \citep{Brysbaert2009}.
These pronunciations are then syllabified, and individual syllables parsed into onset and rime, according to a process described in detail in Appendix \ref{syllabification}.

Wordlikeness ratings from the three studies are plotted against this gross contrast in Figure \ref{boxplot}.
While there are several outliers, there can be little doubt that gross phonotactic status accounts for a considerable amount of variance in wordlikeness judgements.

@@

A wide variety of lexical similarity measures were considered, including a variant of the Generalized Neighborhood model \citep{Bailey2001}, PLD20 \citep{Suarez2011}, and a set of measures provided by the Irvine Phonotactic Calculator \citep{Vaden2009}. 
The measure best correlated with wellformedness judgements is also the most venerable measure of lexical similarity: Coltheart's $N$ \citep{Coltheart1977}, which is defined as the number of words in some representative sample which can be changed into a target nonce word by a single insertion, deletion, or substitution of a phoneme.
\citet{Greenberg1964} find a correlation between wordlikeness ratings and a variant of this measure which only counts words differing by a single substitution.
This is plotted against ratings from the three studies in Figure \ref{neighborhood} with a superimposed local regression (LOESS; \citealt{Cleveland1988}) curve; neighborhood density accounts for much of the variance in ratings.

@@

\begin{figure}[t]
\centering
\includegraphics{neighborhood.pdf}
\caption{Correlation between Coltheart's $N$ and item-averaged wordlikeness ratings, with LOESS curve}
\label{neighborhood}
\end{figure}

While there is nothing inherently ``phonotactic'' about Coltheart's $N$, it indirectly incorporates much of the information present in the gross phonotactic baseline. 
Consider [blɪk]: since there is nothing marked about any part of this nonce word, a ``neighbor'' might be found by modifying any phone: e.g., \emph{click}, \emph{brick}, \emph{bloke}, \emph{bliss}. 
However, since [bn] onsets are unattested in English, a neighbor of [bnɪk] must somehow modify this cluster: this leaves only \emph{brick} and \emph{nick}. 
\citet{Bailey2001} and \citet{Frauenfelder1993} note that neighborhood density is also strongly correlated with measures like bigram probability, but it has been argued elsewhere that phonotactic measures and neighborhood density have distinct effects \citep[e.g.,][]{Berent2003,Pitt1998b,Vitevitch1998,Vitevitch1999a}.

@@

Faciliatory effects of bigram probabilities (i.e., shorter latencies) are reported for other nonce word tasks conducted with adults, including single-word shadowing \citep{Vitevitch1997,Vitevitch1998}, same/different judgements \citep{Lipinski2005,Luce2001,Vitevitch1999a,Vitevitch2005}, and lexical decision \citep{Pylkkanen2002a}. 
\citet{Albright2009a} applies bigram probability as a model of wordlikeness judgements. 
The bigram probability of a sequence $ijk$, for instance, is defined as

@@

\citet{Albright2009a} compares two variants of this model, the first operating over segments, the second over sets of features. 
Unfortunately, the latter model is not described in sufficient detail to allow it to be implemented directly, and there is no publicly available implementation.
However, \citeauthor{Albright2009a}'s evaluation, which includes the \citet{Scholes1966} and \citet{Albright2003b} data, finds an advantage for the segment-based model.
In implementing this model, it was found that a slight improvement could be made by preventing any phoneme-to-phoneme transition from having zero probability. 
This is accomplished by adding 1 to the count of every transition, a technique used in natural language processing and known as Laplace, or ``add one'', smoothing.
As can be seen in Table \ref{bigramcomparison}
this results in a slight increase in the correlation between the scores from this model and wellformedness ratings 
This smoothed segmental bigram score is adopted below.
In Figure \ref{bigram}, it is plotted against wordlikeness ratings from three studies.
